Simulation Study with Feature Contribution

### 2021-09-21 PaperDraft.Rmd
adding Pdf for PLOS one with additional updates to sims

Consider the problem of visualizing multivariate (or potentially high-dimensional) data.  In many cases, it is necessary to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be printed on paper. There are many methods that exist to accomplish this task, ranging from tools such as classicial multidimensional scaling (MDS) to more modern tools like t-distributed Stochastic Neighbor Embedding.

@@

# High-dimensional Data Visualization Tools

@@

Classical multidimensional scaling (MDS) is closely related to PCA (an in some cases, equivalent) (Gower, 1966).  MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations - this distance need not simply be Euclidean. Classical multidimensional scaling returns maps in teh same metric as the original distance or dissimilarity matrix.

## Sparse t-SNE

## Cross Validation and Mantel-esque Test

@@

# Simulation

@@

```{r}
CC2015 <- read.csv("https://raw.githubusercontent.com/paulharmongj/Carnegie_SEM/master/data/CC2015data.csv")

@@

t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset. It is often not obvious that such observations are present when viewing a t-SNE map. In the case of other methods, like MDS, influential points change the configuration of the ordination significantly.

>>>>

Consider the problem of visualizing multivariate (or potentially high-dimensional) data.  In many cases, it is useful to generate a 2 or 3-D mapping of some higher-dimensional space, projecting the high-dimensional data into an ordination that can be directly visualized. There are many methods that exist to accomplish this task, ranging from tools such as classicial multidimensional scaling (MDS) to more modern tools like t-distributed Stochastic Neighbor Embedding.

@@

# Tools for High-Dimensional Data Visualization

@@

Classical multidimensional scaling (MDS) is closely related to PCA (an in some cases, equivalent) (Gower, 1966).  MDS takes as input a distance or dissimilarity matrix representing the distance between pairs of observations. This distance does not have to simply be Euclidean (although it often is). Classical multidimensional scaling returns maps in the same metric as the original distance or dissimilarity matrix; in general, the more dissimilar two observations are in the high-dimensional space, the farther apart they should be in the resulting MDS ordination.

Multidimensional scaling attempts to minimize a metric called "stress" which can be defined generally as:
$$ S_M(z_1, z_2, ..., z_N) = \sum_{i \neq i'}(d_{i,i'} - \|z_i - z_{i'}\|)^2 $$ 
Different versions the stress function exist for different variations of MDS, including methods like Sammon mapping and Classical MDS, which frames the problem from a similarity perspective:

$$ S_c(z_1, z_2, ..., z_N) = \sum_{i, i'} (s_{i,i'} - (z_i - \bar{z}_,z_{i'} - \bar{z}))^2$$ 
Classical MDS can be expressed equivalently as Principal Component Analysis when distances are Euclidean (Elements, XXXX).

## Additional Mapping Techniques

Placeholder section in case we want to add another mapping technique in.

+ SNE
+ Sammon Mapping
+ Isomap

@@

# Simulation Studies

@@

```{r, eval = FALSE}
CC2015 <- read.csv("https://raw.githubusercontent.com/paulharmongj/Carnegie_SEM/master/data/CC2015data.csv")

@@

The proposed method uses a similar framework as to the tools available in regression in order to identify potentially abberant data points that can impact the shape of maps.  Critically, the choice of ordination technique can impact the efficacy and sensitivity of maps to outliers/influential points.  Some methods, like MDS, are highly responsive to influential points that do not look like the rest of the data. Maps resulting from these methods will typically make such points obvious.

Alternatively, t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset. It is often not obvious that such observations are present when viewing a t-SNE map. In many cases, the influential data points are pushed towards the center of the resulting ordination, making them difficult to identify.

Depending on the use case and problem to be solved, this sensitivity (or lack therof) may be benificial or potentially problematic. In some instances, it may be necessary to create maps that are relatively stable even in the presence of influential points; the results in this paper suggest that t-SNE might be more effective for creating maps in those situations.  However, care should be taken when interpreting results from such maps as there may be substantively different observations that get grouped together in the resulting ordination.

Much like in regression, we recommend taking an iterative approach to dealing with influential points.  It is common practice in regression diagnostics to identify the point with the highest Cook's distance, remove it, and re-fit the model to see the new dispersion of observations' impacts on the model.  A similar approach might be useful when creating high-dimensional data maps using these techniques.

We recommend identifying the observation with the largest pseudo-F statistic, removing it, and re-assessing the PERMANOVA results with the reduced set.  By cleaning out some of the most influential points, the resulting maps should be more reproducible, of higher fidelity, and provide potentially more meaningful results.

### 2021-09-21 Simulation with Single Profile.Rmd
adding Pdf for PLOS one with additional updates to sims

---
title: "Simulation Study Document V4"
author: "Paul Harmon"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

@@

This doc contains simulation studies around the efficacy of a Permanova-based method for identifying influential points in a t-SNE or MDS map.  The document steps through a process of simulating data from 3 different groups (currently, these groups are clearly separated) with *3* influential points simulated on a standardized scale 5-6 standard deviations above the observed data.

The code below outlines the number of simulations run - currently this number is small for computational ease but will be increased (with the use of adaptive Permanova methods).

@@

The below chunk runs the simulations. Each iteration simulates a new (contaminated) dataset, runs the t-SNE and MDS mapping procedure, and tests the consistency of the configurations for each of the *n* holdout points. **This code is not run in this version.**

Several versions of the simulations will be run:

+ 1 group with 3 "influental" points
+ 3 groups with 3 "influential" points with different variances
+ 3 groups with 3 influential points with same variances

+ Sample size increase/decrease (number of influential points relative to sample size too)

## Simulation Chunks

This chunk runs the simulation with slighlty increased variability in the three groups, and a restructuring of the influential points. Rather than simulating as outliers all on the positive side, I simulate one that has a large positive set of outlying points, one with a large negative set of outlying points, and one that simply 'bounces around' the multivariate space and breaks the grouping structure.

```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
perplexity = 10

for(j in 1:n_sims){

@@

  g1 <- rmnorm(n = 10, mean = rep(40, 10), diag(45, 10)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(45, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(40, 30), diag(45, 30)) %>%
    t() %>%
    data.frame()

@@

  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()

  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  names(add_outliers) <- names(yes_structure_scale)

  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))

  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
  plot_data_list[[j]] <- datplot

  #### Run the Permanova mapping (for t-SNE)

  xTSNE <- MultiPermanova(yes_structure_final[,1:10], maptype = "tSNE")
  results_list[[j]] <- xTSNE

@@

}
```

# Results {.tabset}

The simulated datasets can be seen in the tabset below.

## Plot List
```{r}
library(ggpubr)
#ggarrange(plotlist = plot_data_list)
print(plot_data_list)
```

## Summary Metrics

We can start to look at some results below. The code generates the rate of correctly classified influential points for both the MDS (shown first) and t-SNE maps. The MDS correctly classifies the influential point *some* of the time, but the t-SNE method very rarely classifies the influential points as having an impact on the configuration.

```{r}

#print the head of the df lists
df3list %>% head(3)
### Identify the number of CORRECTLY Classified influential points
#extract the p-values from the last 3 from MDS
influential_correct <- function(temp){length(which(tail(temp, 3)$PValues < 0.05))/3}
mds_correct <- lapply(df3list, influential_correct)

# extract the pvalues from the last 3 of the t-SNE
tsne_correct <- lapply(df2list, influential_correct)

dat = tibble(MDS = unlist(mds_correct), TSNE = unlist(tsne_correct))
#plot(dat$MDS, dat$TSNE)

mds_correct %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified")
tsne_correct %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified")

Now we can look at the rate of misclassified non-influential points. These are the 60 'regular' points that are spuriously detected as outliers. Each percentage is calculated as a percentage of the 60 that are NOT small p-values (currenlty < 0.05 without a correction). We want a high percentage of correctly-identified non-influential points (although given the machinery of the hypothesis tests used, we would expect at least some spurious detections).

Both are relatively high percentages (as we would expect) - but the t-SNE hovers right around the expected 0.95 and MDS is higher.

```{r}

### Identify rate of CORRECTLY Classified non-influential points
#extract the p-values from the last 3 from MDS
ni_correct <- function(temp){length(which(head(temp, 60)$PValues >= 0.05))/60}

correct_mds <- lapply(df3list, ni_correct)
#kable(correct_mds, col.names = "Correctly Identified Non-Influential: MDS")

# extract the pvalues from the last 3 of the t-SNE
correct_tsne <- lapply(df2list, ni_correct)
#kable(correct_tsne, col.names = "Correctly Identified Non-Influential: tSNE")

correct_mds %>% unlist() %>% mean() %>% kable(col.names = "MDS Percent Correctly Classified Non-Influential")
correct_tsne %>% unlist() %>% mean() %>% kable(col.names = "t-SNE Percent Correctly Classified Non-Influential")

# Plots {.tabset}

## MDS P Values

Plots are shown below for the MDS .

```{r}
plot_mds_results_list
```

## MDS F Stats

```{r}
plot_mds_fstat_results_list
```

## tSNE

```{r}
plot_tsne_results_list
```

## t-SNE F Stats

```{r}
plot_tsne_fstat_results_list
```

>>>>

---
title: "Simulation Study with Feature Contribution"
author: "Paul Harmon"
date: "9/20/2021"
output: html_document
---

@@

This doc looks at a profile where we have aberrant values only on a few of the features. Can we detect?

@@

## Profile

Let's look at a couple example profiles - here we might have a feature with only a few weird values.

```{r sims_v2, eval = TRUE, cache = TRUE, warning = FALSE, message = FALSE}

@@

  g1 <- rmnorm(n = 20, mean = rep(40, 10), diag(45, 10)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 20, mean = rep(40, 20), diag(45, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 20, mean = rep(40, 30), diag(45, 30)) %>%
    t() %>%
    data.frame()

@@

  yes_structure_scale <- lapply(yes_structure[,1:20], scale) %>% as_tibble()

  ### adds outlier - where it's roughly the same for all but 2 of the features
  add_outliers <- c(rnorm(18, 0, 1), c(-5,5))
  names(add_outliers) <- names(yes_structure_scale)

  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))

  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:20) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
  datplot

```{r}
perplexity = 10
  #### Run the Permanova mapping (for t-SNE)

  xTSNE <- MultiPermanova(yes_structure_final[,1:20], maptype = "tSNE")
  results_list[[j]] <- xTSNE

@@

}
